---
title: "01 - Optimal consumption and DWL under current pricing"
author: "Nathan DeMaagd"
date: "9/29/2020"
output: html_document
---




This markdown file estimates a demand function for each household using elasticity estimates from (DeMaagd and Roberts(2020))[https://www.researchgate.net/publication/343825978_Estimating_water_demand_using_price_differences_of_wastewater_services], and calculates optimal consumption with and without consideration of marginal user cost. It then calculates deadweight loss. It is dependent on the scripts in the 'MAIN/Functions' folder and parameters defined in the 'MAIN/00-definitions.R' file.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# set wd
knitr::opts_knit$set(root.dir = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/WaterMarginalAveragePrice/MAIN/')


# load packages
library(doParallel); library(ggplot2)

registerDoParallel(cores = 10)

```




Load water use data and parameters. NOTE: Estimate of MUC comes from (Ordering the extraction of renewable resources: The case of multiple aquifers (Roumasset and Wada, 2012))[https://www.sciencedirect.com/science/article/abs/pii/S0928765511000625]. They estimate an efficiency price of \$2.16/tgal and a marginal price of \$0.31/tgal, giving a MUC of \$1.84/tgal.

```{r load data and parameters, results = 'hide'}


# load definitions
source("00-definitions.R")


# load billing data
load("C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Cesspools/Data/regression_dat_FINAL.rdata")

```




# 00 - Calculate consumer bills

First, we use the pricing information to calculate average monthly bills, given households' average monthly water usage.

```{r monthly bills}


# create 30-day monthly consumption value
all.sfds$avg.30day.use.1000s <- all.sfds$avg.daily.use * 30 / 1000
all.sfds$avg.30day.use       <- all.sfds$avg.daily.use * 30


# calculate bill (just water, no sewer)
                                    # find block                                                                                       # add appropriate charges for consumption level
all.sfds$total.water.bill <- ifelse(all.sfds$avg.30day.use <= volume_water_blk1_qty,                                                   fixed_water_rate + volume_water_blk1_rate * all.sfds$avg.30day.use,
                                    
                             ifelse(all.sfds$avg.30day.use >  volume_water_blk1_qty & all.sfds$avg.30day.use <= volume_water_blk2_qty, fixed_water_rate + volume_water_blk1_rate * volume_water_blk1_qty +
                                                                                                                                           volume_water_blk2_rate * (all.sfds$avg.30day.use - volume_water_blk1_qty),
                                           
                             ifelse(all.sfds$avg.30day.use >  volume_water_blk2_qty,                                                   fixed_water_rate + volume_water_blk1_rate * volume_water_blk1_qty +
                                                                                                                                           volume_water_blk2_rate * volume_water_blk2_qty +
                                                                                                                                           volume_water_blk3_rate * (all.sfds$avg.30day.use - volume_water_blk2_qty), NA)))


# if household has sewer service, add on the fixed and volumetric charge (BWS calculates sewer bill with 80% of HH's consumption to account for irrigation)
all.sfds$total.water.sewer.bill <- ifelse(all.sfds$no.sewer == 'Sewer', all.sfds$total.water.bill + fixed_sewer_rate + volume_sewer_rate * all.sfds$avg.30day.use * 0.80, all.sfds$total.water.bill)


# calculate average cost per gal
all.sfds$total.water.sewer.bill.avg.per.gal <- all.sfds$total.water.sewer.bill / all.sfds$avg.30day.use

```




# 01 - Demand functions

First we use current household consumption to estimate a demand curve, assuming the functional form P = AQ^(1/e). We use price elasticity demand estimates from the [water demand](https://www.researchgate.net/publication/343825978_Estimating_water_demand_using_price_differences_of_wastewater_services) paper. The elasticity and its confidence interval is used to estimate the $A$ parameter.

```{r 01 - demand function estimate}


# source the demand curve function
source('Code/Functions/01-demand-curve-calculation.R')


# for the different elasticity estimates, calculate A = P / Q^(1/e)
all.sfds$A_elasticity020 <- demand_A(bill_avg_per_gal = all.sfds$total.water.sewer.bill.avg.per.gal,
                                     avg_monthly_gal  = all.sfds$avg.30day.use,
                                     elasticity       = -0.020)

all.sfds$A_elasticity060 <- demand_A(bill_avg_per_gal = all.sfds$total.water.sewer.bill.avg.per.gal,
                                     avg_monthly_gal  = all.sfds$avg.30day.use,
                                     elasticity       = -0.060)

all.sfds$A_elasticity326 <- demand_A(bill_avg_per_gal = all.sfds$total.water.sewer.bill.avg.per.gal,
                                     avg_monthly_gal  = all.sfds$avg.30day.use,
                                     elasticity       = -0.326)


# get median A by group
avg_A_elasticity_020 <- aggregate(all.sfds$A_elasticity020, list(all.sfds$no.sewer), median)
avg_A_elasticity_060 <- aggregate(all.sfds$A_elasticity060, list(all.sfds$no.sewer), median)
avg_A_elasticity_326 <- aggregate(all.sfds$A_elasticity326, list(all.sfds$no.sewer), median)


# save `avg_A` files for use in figures
saveRDS(avg_A_elasticity_020, file = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01b-avg_demand_A_vals_elasticity020.rds')
saveRDS(avg_A_elasticity_060, file = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01b-avg_demand_A_vals_elasticity060.rds')
saveRDS(avg_A_elasticity_326, file = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01b-avg_demand_A_vals_elasticity326.rds')


rm(demand_A, avg_A_elasticity_020, avg_A_elasticity_060, avg_A_elasticity_326)

```




# 02 - Optimal consumption

With the demand curves estimated for each household, we can calculate the present-day utility-maximizing and socially-optimal levels of consumption for each household. Utility-maximizing means present-day utility, with no regard to the marginal user cost of water use. Socially-optimal means consideration for MUC; we test an estimate from (Ordering the extraction of renewable resources: The case of multiple aquifers (Roumasset and Wada, 2012)[https://www.sciencedirect.com/science/article/abs/pii/S0928765511000625].

```{r 02 - optimal consumption}


# load optimal consumption function
source('Code/Functions/02-uniroot-optimal-consumption.R')



### optimal consumption: elasticity -0.020

elasticity_demand = -0.020

# run for MUC = $0, MUC = $1.84/tgal, and MUC = $3.68/tgal
all.sfds$avg.30day.use_optimal_elasticity020_MUC000 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity020, MUC = 0)
all.sfds$avg.30day.use_optimal_elasticity020_MUC184 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity020, MUC = MUC)
all.sfds$avg.30day.use_optimal_elasticity020_MUC368 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity020, MUC = MUC*2)



### optimal consumption: elasticity -0.060

elasticity_demand = -0.060

# run for MUC = $0, MUC = $1.84/tgal, and MUC = $3.68/tgal
all.sfds$avg.30day.use_optimal_elasticity060_MUC000 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity060, MUC = 0)
all.sfds$avg.30day.use_optimal_elasticity060_MUC184 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity060, MUC = MUC)
all.sfds$avg.30day.use_optimal_elasticity060_MUC368 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity060, MUC = MUC*2)



### optimal consumption: elasticity -0.326

elasticity_demand = -0.326

# run for MUC = $0, MUC = $1.84/tgal, and MUC = $3.68/tgal
all.sfds$avg.30day.use_optimal_elasticity326_MUC000 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity326, MUC = 0)
all.sfds$avg.30day.use_optimal_elasticity326_MUC184 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity326, MUC = MUC)
all.sfds$avg.30day.use_optimal_elasticity326_MUC368 <- optimal_consumption(data = all.sfds, A = all.sfds$A_elasticity326, MUC = MUC*2)


# return elasticity to original value
elasticity_demand = -0.060




### optimal consumption: elasticity -0.060 with consumer quantity uncertainty and no MUC

# run bootstrap
consumption_bootstrap <- foreach(i = 1:5000, .combine = 'cbind') %dopar% {

  all.sfds$avg.30day.use_optimal_elasticity060_MUC000 * (runif(nrow(all.sfds), min = 0.8, max = 1.2))

}

# get rowmeans for boostrapped consumption with uncertainty
all.sfds$avg.30day.use_optimal_elasticity060_MUC000_uncertainty20pct <- rowMeans(consumption_bootstrap)









##### plot distributions #####




### box and whisker

# create box and whisker plots comparing optimal water use for mean elasticity estimate (-0.06)

# melt data
water_use_melt <- data.frame(value    = c(all.sfds$avg.30day.use, all.sfds$avg.30day.use_optimal_elasticity060_MUC000, all.sfds$avg.30day.use_optimal_elasticity060_MUC184,
                                          all.sfds$avg.30day.use_optimal_elasticity060_MUC368),
                             variable = rep(factor(c('Current use', 'Present-day\nutility maximizing', 'Socially optimal,\nMUC = $1.84/tgal', 'Socially optimal,\nMUC = $3.68/tgal'),
                                                   levels = c('Current use', 'Present-day\nutility maximizing', 'Socially optimal,\nMUC = $1.84/tgal', 'Socially optimal,\nMUC = $3.68/tgal')),
                                            each = nrow(all.sfds)))


# plot distributions of water use: current, utility-maximizing, MUC = 1.84, and MUC = 3.68
water_use_distributions <- ggplot(data = water_use_melt) + geom_boxplot(aes(x = variable, y = value/1000),
                                             outlier.shape = NA) +
  
  # change axis labels
  labs(x = 'Consumption type', y = 'Household consumption (tgal/month)') +
  
  # adjust plot appearance
  theme(legend.title = element_blank(), text = element_text(size = 15),
        axis.ticks = element_blank(),
        panel.grid.major.x = element_blank()) +
  
  # adjust y scale to omit outliers
  coord_cartesian(ylim = quantile(water_use_melt$value/1000, c(0, 0.95), na.rm = TRUE)) +
  
  # add median use to median bar in boxplots
  annotate("text", x = 1:4,
           y     = c(             median(water_use_melt[water_use_melt$variable == 'Current use',                         1], na.rm = TRUE)/1000+0.5,
                                  median(water_use_melt[water_use_melt$variable == 'Present-day\nutility maximizing',     1], na.rm = TRUE)/1000+0.5,
                                  median(water_use_melt[water_use_melt$variable == 'Socially optimal,\nMUC = $1.84/tgal', 1], na.rm = TRUE)/1000+0.5,
                                  median(water_use_melt[water_use_melt$variable == 'Socially optimal,\nMUC = $3.68/tgal', 1], na.rm = TRUE)/1000+0.5),
           
           label = c(format(round(median(water_use_melt[water_use_melt$variable == 'Current use',                         1], na.rm = TRUE)/1000, 2), nsmall = 2),
                     format(round(median(water_use_melt[water_use_melt$variable == 'Present-day\nutility maximizing',     1], na.rm = TRUE)/1000, 2), nsmall = 2),
                     format(round(median(water_use_melt[water_use_melt$variable == 'Socially optimal,\nMUC = $1.84/tgal', 1], na.rm = TRUE)/1000, 2), nsmall = 2),
                     format(round(median(water_use_melt[water_use_melt$variable == 'Socially optimal,\nMUC = $3.68/tgal', 1], na.rm = TRUE)/1000, 2), nsmall = 2)))

ggsave(plot = water_use_distributions, filename = 'Figures/Images/02-optimal-water-use-distributions.png', height = 5, width = 7, dpi = 300)




### histograms - current, utility-maximizing, socially-optimal, uncertainty in consumption


# current consumption
hist_current <- ggplot(data = all.sfds) +
  geom_histogram(aes(avg.30day.use/1000), bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = c(0, 40)) +
  labs(x = 'Mean household water use (1000s gal/month)', y = 'Number of households') +
  geom_vline(xintercept = c(13, 30), linetype = 'longdash') +
  theme(text = element_text(size = 15))

ggsave(plot = hist_current, filename = 'Figures/Images/01-hist_current-water-use.png', height = 5, width = 8, dpi = 300)


# optimal consumption, elasticity -0.06 and no MUC
hist_utilmax_elasticity060 <- ggplot(data = all.sfds) +
  geom_histogram(aes(avg.30day.use_optimal_elasticity060_MUC000/1000), bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = c(0, 40)) +
  labs(x = 'Present-day utility maximizing household water use (1000s gal/month)', y = 'Number of households') +
  geom_vline(xintercept = c(13, 30), linetype = 'longdash') +
  theme(text = element_text(size = 15))

ggsave(plot = hist_utilmax_elasticity060, filename = 'Figures/Images/01-hist-utilmax-elasticity060-MUC000-water-use.png', height = 5, width = 8, dpi = 300)


# optimal consumption, elasticity -0.06 and MUC 1.84
hist_socialmax_elasticity060 <- ggplot(data = all.sfds) +
  geom_histogram(aes(avg.30day.use_optimal_elasticity060_MUC184/1000), bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = c(0, 40)) +
  labs(x = 'Socially-optimal household water use (1000s gal/month)', y = 'Number of households') +
  geom_vline(xintercept = c(13, 30), linetype = 'longdash') +
  theme(text = element_text(size = 15))

ggsave(plot = hist_socialmax_elasticity060, filename = 'Figures/Images/01-hist-socialmax-elasticity006-MUC184-water-use.png', height = 5, width = 8, dpi = 300)


# optimal consumption, elasticity -0.06, MUC 000, and uncertainty in consumption
hist_utilmax_elasticity060_uncertainty <- ggplot(data = all.sfds) +
  geom_histogram(aes(avg.30day.use_optimal_elasticity060_MUC000_uncertainty20pct/1000), bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = c(0, 40)) +
  labs(x = 'Present-day utility maximizing household water use (1000s gal/month)', y = 'Number of households') +
  geom_vline(xintercept = c(13, 30), linetype = 'longdash') +
  theme(text = element_text(size = 15))

ggsave(plot = hist_utilmax_elasticity060_uncertainty, filename = 'Figures/Images/01-hist-utilmax-elasticity060-MUC000-uncertainty-water-use.png', height = 5, width = 8, dpi = 300)


rm(optimal_consumption, Q_intercept_fcn, water_use_melt)

```




# 03 - deadweight loss

We now know the difference between current and optimal consumption (with and without consideration of scarcity pricing). Use these values, the demand curves, and the pricing function to estimate monthly deadweight loss for each household.

```{r 03 - deadweight loss}


# load deadweight loss function
source('Code/Functions/03-deadweight-loss-fcn.R')




##### household-level DWL #####

# find DWL for each combination of elasticity and MUC value


# elasticity -0.020; MUC = 0, MUC = 1.84, and MUC = 3.68

elasticity_demand <- -0.020

  # if consumers respond to avg price
  all.sfds$DWL_avp_elasticity020_MUC184 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity020_MUC184[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity020_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity020[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_avp_elasticity020_MUC368 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity020_MUC368[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity020_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity020[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  # if consumers respond to marginal price
  all.sfds$DWL_mp_elasticity020_MUC184  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity020_MUC184[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity020_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity020_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity020[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_mp_elasticity020_MUC368  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity020_MUC368[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity020_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity020_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity020[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  
# elasticity -0.060; MUC = 0, MUC = 1.84, and MUC = 3.68

elasticity_demand <- -0.060

  # if consumers respond to avg price
  all.sfds$DWL_avp_elasticity060_MUC184 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity060_MUC184[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity060_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity060[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_avp_elasticity060_MUC368 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity060_MUC368[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity060_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity060[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  # if consumers respond to marginal price
  all.sfds$DWL_mp_elasticity060_MUC184  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity060_MUC184[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity060_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity060_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity060[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_mp_elasticity060_MUC368  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity060_MUC368[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity060_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity060_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity060[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  
# elasticity -0.326; MUC = 0, MUC = 1.84, and MUC = 3.68

elasticity_demand <- -0.326

  # if consumers respond to avg price
  all.sfds$DWL_avp_elasticity326_MUC184 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity326_MUC184[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity326_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity326[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_avp_elasticity326_MUC368 <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_avp_elasticity326_MUC368[[i]] <- deadweight_loss(q_actual  = all.sfds$avg.30day.use[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity326_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity326[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  # if consumers respond to marginal price
  all.sfds$DWL_mp_elasticity326_MUC184  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity326_MUC184[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity326_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity326_MUC184[[i]],
                                                                                           A = all.sfds$A_elasticity326[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  all.sfds$DWL_mp_elasticity326_MUC368  <- NA
  for(i in 1:nrow(all.sfds)){all.sfds$DWL_mp_elasticity326_MUC368[[i]]  <- deadweight_loss(q_actual  = all.sfds$avg.30day.use_optimal_elasticity326_MUC000[[i]],
                                                                                           q_optimal = all.sfds$avg.30day.use_optimal_elasticity326_MUC368[[i]],
                                                                                           A = all.sfds$A_elasticity326[[i]], no.sewer = all.sfds$no.sewer[[i]])}
  
  
  
  
##### island-wide SFD DWL #####
  
# calculate the total DWL of all single family homes on the island

sum(all.sfds$DWL_avp_elasticity020_MUC184, na.rm = TRUE)
sum(all.sfds$DWL_avp_elasticity020_MUC368, na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity020_MUC184,  na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity020_MUC368,  na.rm = TRUE)

sum(all.sfds$DWL_avp_elasticity060_MUC184, na.rm = TRUE)
sum(all.sfds$DWL_avp_elasticity060_MUC368, na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity060_MUC184,  na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity060_MUC368,  na.rm = TRUE)

sum(all.sfds$DWL_avp_elasticity326_MUC184, na.rm = TRUE)
sum(all.sfds$DWL_avp_elasticity326_MUC368, na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity326_MUC184,  na.rm = TRUE)
sum(all.sfds$DWL_mp_elasticity326_MUC368,  na.rm = TRUE)

```




# save data

```{r save data}


saveRDS(all.sfds, file = "C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01_RESULTS_optimal_consumption_DWL_current_pricing.rds")

```