---
title: "02 - Revenue analysis under alternative pricing schemes"
author: "Nathan DeMaagd"
date: "9/30/2020"
output: html_document
---




This markdown file calculates current utility revenue. It then compares revenue and relative efficiency of alternative pricing schemes. It is dependent on the scripts in the 'MAIN/Functions' folder and parameters defined in the 'MAIN/00-definitions.R' file.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# set wd
knitr::opts_knit$set(root.dir = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/WaterMarginalAveragePrice/MAIN/')


# load packages
library(doParallel); library(ggplot2)

registerDoParallel(cores = 10)

```




Load water use data and parameters. NOTE: Estimate of MUC comes from (Ordering the extraction of renewable resources: The case of multiple aquifers (Roumasset and Wada, 2012))[https://www.sciencedirect.com/science/article/abs/pii/S0928765511000625]. They estimate an efficiency price of \$2.16/tgal and a marginal price of \$0.31/tgal, giving a MUC of \$1.84/tgal.

```{r load data and parameters, results = 'hide'}


# load definitions
source("00-definitions.R")


# load billing data with optimal consumption and DWL from previous Rmd file
all.sfds <- readRDS("C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01_RESULTS_optimal_consumption_DWL_current_pricing.rds")

```




# 04 - First analyze revenue under current pricing

Under the current block prices, what are consumers paying for their water? What is the total revenue earned by the utility each month?

```{r 04 - revenue under current pricing}


# total revenue
total_revenue_current       <- sum(all.sfds$total.water.bill, na.rm = TRUE)/1e6
total_revenue_current_label <- paste0('$', round(total_revenue_current, 2))


# boxplot of water bill (note we exclude the sewer portion since that is not revenue for the water utility)
current_revenue_density <- ggplot(data = all.sfds) +
  geom_histogram(aes(total.water.bill),
                 bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = quantile(all.sfds$total.water.bill, probs = c(0.01, 0.98))) +
  labs(x = 'Household average water bill ($/mo)', y = 'Number of households') +
  annotate('text', x = 100, y = 4000, label = paste0("Water utility current revenue\n", total_revenue_current_label, ' mil/mo'), hjust = 0)

ggsave(filename = "Figures/Images/04-current-water-bills.png", height = 4, width = 7)

```




# 05 - Constant marginal price, no fixed price

Using household demand functions, find a marginal price that would yield the same revenue as the current pricing structure, assuming no fixed charge. 

```{r constant marginal, no fixed}


# load demand function
source('Code/Functions/05-demand-given-price.R')




##### first test marginal rate = avg $/gal of current system #####


# get summary of current $/gal for each household
  # no sewer: current mean is $0.008194904 per gallon; median is $0.005546433 per gallon
  # sewer:    current mean is $0.02566547  per gallon; median is $0.02007204  per gallon
aggregate(all.sfds$total.water.sewer.bill.avg.per.gal, list(all.sfds$no.sewer), summary)


# calculate monthly water use under Constant Marginal Price (based on mean and median of current average prices)
all.sfds$avg.30day.use_CMPmeancurrent_elasticity020   <- demand(A = all.sfds$A_elasticity020, price_water = 0.008194904, price_water_sewer = 0.02566547, elasticity = -0.020, no.sewer = all.sfds$no.sewer)
all.sfds$avg.30day.use_CMPmeancurrent_elasticity060   <- demand(A = all.sfds$A_elasticity060, price_water = 0.008194904, price_water_sewer = 0.02566547, elasticity = -0.060, no.sewer = all.sfds$no.sewer)
all.sfds$avg.30day.use_CMPmeancurrent_elasticity326   <- demand(A = all.sfds$A_elasticity326, price_water = 0.008194904, price_water_sewer = 0.02566547, elasticity = -0.326, no.sewer = all.sfds$no.sewer)


# calculate revenue under this pricing
total_revenue_CMPmeancurrent_elasticity020 <- all.sfds$avg.30day.use_CMPmeancurrent_elasticity020 * 0.008194904
total_revenue_CMPmeancurrent_elasticity020 <- sum(total_revenue_CMPmeancurrent_elasticity020[!is.infinite(total_revenue_CMPmeancurrent_elasticity020)], na.rm = TRUE) / 1e6

total_revenue_CMPmeancurrent_elasticity060 <- all.sfds$avg.30day.use_CMPmeancurrent_elasticity060 * 0.008194904
total_revenue_CMPmeancurrent_elasticity060 <- sum(total_revenue_CMPmeancurrent_elasticity060[!is.infinite(total_revenue_CMPmeancurrent_elasticity060)], na.rm = TRUE) / 1e6

total_revenue_CMPmeancurrent_elasticity326 <- all.sfds$avg.30day.use_CMPmeancurrent_elasticity326 * 0.008194904
total_revenue_CMPmeancurrent_elasticity326 <- sum(total_revenue_CMPmeancurrent_elasticity326[!is.infinite(total_revenue_CMPmeancurrent_elasticity326)], na.rm = TRUE) / 1e6




# plot consumption under this pricing scheme


    # melt data
    all.sfds.melt <- data.frame(value = c(all.sfds$avg.30day.use, all.sfds$avg.30day.use_CMPmeancurrent_elasticity020, all.sfds$avg.30day.use_CMPmeancurrent_elasticity060,
                                          all.sfds$avg.30day.use_CMPmeancurrent_elasticity326),
                                variable = rep(c('Current', 'Elasticity = -0.020', 'Elasticity = -0.060', 'Elasticity = -0.326'), each = nrow(all.sfds)))
    
    
    # get median values by group to add to boxplot
    cons_by_group <- aggregate(all.sfds.melt$value, by = list(all.sfds.melt$variable), median)
    
    
CMP_consumption_distributions <- ggplot(data = all.sfds.melt) +
  
  geom_boxplot(aes(x = variable, y = value), outlier.shape = NA) +
  
  scale_y_continuous(limits = quantile(all.sfds.melt$value, probs = c(0, 0.90))) +
  
  scale_x_discrete(labels = c('Current', 'Elasticity = -0.020', 'Elasticity = -0.060', 'Elasticity = -0.326')) +
  
  labs(x = NULL, y = 'Gallons per month per household') +
  
  annotate('text', x = 1:4, y = (cons_by_group$x)+50, label = round(cons_by_group$x)) +
  
  theme(legend.title = element_blank())


ggsave(plot = CMP_consumption_distributions, filename = "Figures/Images/05-water-consumption-under-CMPcurrentmean.png", height = 5, width = 7)




##### second: test marginal rate = marginal cost + MUC #####


# define marginal cost
MC = 0.31 / 1000


# calculate monthly water use under Constant Marginal Price (based on marginal cost + MUC, average sewer price)
all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity020   <- demand(A = all.sfds$A_elasticity020, price_water = MC + MUC, price_water_sewer = 0.02566547, elasticity = -0.020, no.sewer = all.sfds$no.sewer)
all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity060   <- demand(A = all.sfds$A_elasticity060, price_water = MC + MUC, price_water_sewer = 0.02566547, elasticity = -0.060, no.sewer = all.sfds$no.sewer)
all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity326   <- demand(A = all.sfds$A_elasticity326, price_water = MC + MUC, price_water_sewer = 0.02566547, elasticity = -0.326, no.sewer = all.sfds$no.sewer)


# calculate revenue under this pricing
total_revenue_CMP_MC_MUC184_elasticity020 <- all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity020 * (MC + MUC)
total_revenue_CMP_MC_MUC184_elasticity020 <- sum(total_revenue_CMP_MC_MUC184_elasticity020[!is.infinite(total_revenue_CMP_MC_MUC184_elasticity020)], na.rm = TRUE) / 1e6

total_revenue_CMP_MC_MUC184_elasticity060 <- all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity060 * (MC + MUC)
total_revenue_CMP_MC_MUC184_elasticity060 <- sum(total_revenue_CMP_MC_MUC184_elasticity060[!is.infinite(total_revenue_CMP_MC_MUC184_elasticity060)], na.rm = TRUE) / 1e6

total_revenue_CMP_MC_MUC184_elasticity326 <- all.sfds$avg.30day.use_CMP_MC_MUC184_elasticity326 * (MC + MUC)
total_revenue_CMP_MC_MUC184_elasticity326 <- sum(total_revenue_CMP_MC_MUC184_elasticity326[!is.infinite(total_revenue_CMP_MC_MUC184_elasticity326)], na.rm = TRUE) / 1e6


```
