---
title: "02 - Revenue analysis under alternative pricing schemes"
author: "Nathan DeMaagd"
date: "9/30/2020"
output: html_document
---




This markdown file calculates current utility revenue. It then compares revenue and relative efficiency of alternative pricing schemes. It is dependent on the scripts in the 'MAIN/Functions' folder and parameters defined in the 'MAIN/00-definitions.R' file.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# set wd
knitr::opts_knit$set(root.dir = 'C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/WaterMarginalAveragePrice/MAIN/')


# load packages
library(doParallel); library(ggplot2)

registerDoParallel(cores = 10)

```




Load water use data and parameters. NOTE: Estimate of MUC comes from (Ordering the extraction of renewable resources: The case of multiple aquifers (Roumasset and Wada, 2012))[https://www.sciencedirect.com/science/article/abs/pii/S0928765511000625]. They estimate an efficiency price of \$2.16/tgal and a marginal price of \$0.31/tgal, giving a MUC of \$1.84/tgal.

```{r load data and parameters, results = 'hide'}


# load definitions
source("00-definitions.R")


# load billing data with optimal consumption and DWL from previous Rmd file
all.sfds <- readRDS("C:/Users/nated/OneDrive - hawaii.edu/Documents/Projects/Water/Marginal vs average price welfare/Data/Intermediate/01_RESULTS_optimal_consumption_DWL_current_pricing.rds")

```




# 04 - First analyze revenue under current pricing

Under the current block prices, what are consumers paying for their water? What is the total revenue earned by the utility each month?

```{r 04 - revenue under current pricing}


# total revenue
total_revenue_current     <- sum(all.sfds$total.water.bill, na.rm = TRUE)
total_revenue_label       <- paste0('$', round(total_revenue/1e6, 2))


# boxplot of water bill (note we exclude the sewer portion since that is not revenue for the water utility)
current_revenue_density <- ggplot(data = all.sfds) +
  geom_histogram(aes(total.water.bill),
                 bins = 100, alpha = 0.6, color = 'gray30', fill = 'gray30') +
  scale_x_continuous(limits = quantile(all.sfds$total.water.bill, probs = c(0.01, 0.98))) +
  labs(x = 'Household average water bill ($/mo)', y = 'Number of households') +
  annotate('text', x = 100, y = 4000, label = paste0("Water utility current revenue\n", total_revenue_mil, ' mil/mo'), hjust = 0)

ggsave(filename = "Figures/Images/04-current-water-bills.png", height = 4, width = 7)

```
